### 2012-03-13 docs/notes.tex
Add gradient derivation.  Add material/figs to slides.

 \centerline{\Large \bf Stochastic Blockmodels for Relational Events} 
 \medskip
\centerline{\bf Christopher DuBois}
 \bigskip

\abstract{

Stochastic blockmodels are a tool for exploring the structure of static social networks by learning latent groups of individuals who have similar tendencies in their  group-wise interactions. Continuous-time methods, such as the relational event framework (Butts 2008), model the rate of dyadic events conditioned on the observed history of events and covariates about the actors.  We propose a hierarchical model combining these two approaches where each pair of blocks has a set of parameters guiding the event dynamics for dyads between those blocks.  The  method is illustrated using  dyadic interaction data (e.g. email, Twitter) among hundreds of actors and thousands of events.  We show parameter estimates from the model can reveal heterogeneity in the dynamics among groups of individuals, providing a better fit to the data than a single relational event model for all actors in the network.  Our approach illustrates the efficacy of combining a detailed model for local dependencies and a latent variable model for meso-scale dependencies.

@@

\section{Simulation}
\begin{figure}
\center
\includegraphics[width=2in]{../figs/syn/mat.pdf}
\includegraphics[width=3in]{../figs/syn/counts.pdf}
\caption{Illustration of 1000 simulated events, as described in text. Left: Counts of each dyad. Right: Boxplot of distribution of participation counts across dyads.  The top left shows an increased propensity for reciprocity within cluster 1; bottom right shows more AB-AY events within cluster 2.}
\label{fig:syncounts}
\end{figure}

@@

\begin{figure}
\center
\includegraphics[width=3in]{../figs/syn/logposterior.pdf}
\includegraphics[width=3in]{../figs/syn/test-recall.pdf}
\caption{Results from fitting models to synthetic data. Left: Log posterior of 1000 simulated events vs. iteration during MCMC.  Right: Recall on 3000 test events.}
\label{fig:synresults}
\end{figure}
We evaluate the predictive ability of the learned models by comparing models based on the loglikelihood of held-out data and recall on held out data.  Each data set is first split into a training set and a test set, and the loglikelihood of the test set is computed using Equation \ref{eqn:llk} where $\beta$ is set to be the mean from posterior samples given the training data.

@@

I do not believe it is possible to perform EM for a model where we have the more general linear predictor for the intensity function.  Since the assignment of one node depends on the assignments of the other nodes, we suffer combinatorial explosion.
\bibliographystyle{plain}
\bibliography{refs}

>>>>

 \centerline{\Large \bf Stochastic Blockmodels for Relational Event Data} 
 \medskip
\centerline{\bf Christopher DuBois}
 \bigskip

\abstract{
Stochastic blockmodels assume latent groups of individuals have similar tendencies in their  group-wise interactions.  Many find these methods useful for exploratory analysis of static networks. On the other hand, continuous-time methods such as the relational event framework (Butts 2008) model the rate of dyadic events conditioned on the observed history of events and covariates about the actors.  We propose to combine these two approaches where each pair of blocks has a set of parameters guiding the event dynamics for dyads between those blocks.  The method is illustrated using dyadic interaction data (e.g. email, Twitter direct messages) among hundreds of actors and thousands of events.  We show parameter estimates from the model can reveal heterogeneity in the dynamics among groups of individuals, providing a better fit to the data than using a single relational event model.  Our approach illustrates the efficacy of combining a detailed model for local dependencies and a latent variable model for meso-scale dependencies.

%Stochastic blockmodels are a tool for exploring the structure of static social networks by learning latent groups of individuals who have similar tendencies in their  group-wise interactions. Continuous-time methods, such as the relational event framework (Butts 2008), model the rate of dyadic events conditioned on the observed history of events and covariates about the actors.  We propose a hierarchical model combining these two approaches where each pair of blocks has a set of parameters guiding the event dynamics for dyads between those blocks.  The  method is illustrated using  dyadic interaction data (e.g. email, Twitter) among hundreds of actors and thousands of events.  We show parameter estimates from the model can reveal heterogeneity in the dynamics among groups of individuals, providing a better fit to the data than a single relational event model for all actors in the network.  Our approach illustrates the efficacy of combining a detailed model for local dependencies and a latent variable model for meso-scale dependencies.

@@

\section{Simulation}
\begin{figure}
\center
\includegraphics[width=2in]{../figs/synthetic/mat.pdf}
\includegraphics[width=3in]{../figs/synthetic/counts.pdf}
\caption{Illustration of 1000 simulated events, as described in text. Left: Counts of each dyad. Right: Boxplot of distribution of participation counts across dyads.  The top left shows an increased propensity for reciprocity within cluster 1; bottom right shows more AB-AY events within cluster 2.}
\label{fig:syncounts}
\end{figure}

@@

\begin{figure}
\center
%\includegraphics[width=3in]{../figs/synthetic/logposterior.pdf}
\includegraphics[width=3in]{../figs/synthetic/test-recall.pdf}
\caption{Results from fitting models to synthetic data. Left: Log posterior of 1000 simulated events vs. iteration during MCMC.  Right: Recall on 3000 test events.}
\label{fig:synresults}
\end{figure}
We evaluate the predictive ability of the learned models by comparing models based on the loglikelihood of held-out data and recall on held out data.  Each data set is first split into a training set and a test set, and the loglikelihood of the test set is computed using Equation \ref{eqn:llk} where $\beta$ is set to be the mean from posterior samples given the training data.

@@

I do not believe it is possible to perform EM for a model where we have the more general linear predictor for the intensity function.  Since the assignment of one node depends on the assignments of the other nodes, we suffer combinatorial explosion.

\section{Gradient}

\begin{align*}
\frac{\partial \ell}{\partial \beta_{p,k,l}} = \frac{\partial}{\partial \beta_{p,k,l}} \sum_{m = 1}^M \left[ \log \lambda_{i_m,j_m}(t_m | \cdot) - \sum_{(i,j) \in \mathcal{R}_{i_m,j_m}} (t_m-\tau_{i,j,m}) \lambda_{i,j}(t_m|\cdot) \right] \\ 
  =\sum_{m \in V} \left[ s_p(t_m,i_m,j_m) - \sum_{(i,j) \in \mathcal{R}_{i_m,j_m}} (t_m-\tau_{i,j,m}) s_p(t_m,i,j) \lambda_{i,j}(t_m|\cdot) \right]
\end{align*}

where $V$ is the set of $m$ where either $z_i \in \{k,l\}$ or $z_j \in \{k,l\}$.

\bibliographystyle{plain}
\bibliography{refs}

### 2012-03-13 docs/slides/slides.tex
Add gradient derivation.  Add material/figs to slides.

\documentclass{beamer}
\usetheme{default}
\usepackage{tikz}
\usepackage{subfigure}
\tikzstyle{vertex}=[auto=left,circle,fill=black!25,minimum size=20pt,inner sep=0pt]
\title[Blockmodels for relational events]%(optional, only for long titles)
{Stochastic blockmodels for relational events}
\subtitle{}
\author[DuBois, Christopher] % (optional, for multiple authors)
{Christopher~DuBois\inst{1} \and Carter~Butts\inst{2} \and Padhraic~Smyth\inst{3}}
\institute[] % (optional)
{
  \inst{1}%
Department of Statistics \\
University of California, Irvine
  \and
  \inst{2}%
Department of Sociology \\
University of California, Irvine
  \and
  \inst{2}%
Department of Computer Science \\
University of California, Irvine
}
\date[Sunbelt 2012] % (optional)
{Sunbelt 2012}
\subject{Informatik}
\begin{document}
\frame{\titlepage}

@@

\begin{align}
\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}_{i_m,j_m}}\exp\{ - (t_m - \tau_{i,j,m}) \lambda_{ij}(t_m | .) \}
\end{align}

\begin{itemize}
\item $t_m$: time of event $m$
\item $\tau_{i,j,m}$: changepoint time $\lambda_{ij}(t)$ prior to $t_m$
\item $\mathcal{R}_{i,j}$: set of dyads involving either $i$ or $j$
\end{itemize}

@@

\includegraphics[scale=.5]{../../figs/synthetic/unsorted}
\includegraphics[scale=.5]{../../figs/synthetic/mat}
\includegraphics[scale=.5]{../../figs/synthetic/bm}

\vspace{-.8cm}
\begin{align*}
\mbox{Mixing rates} & & \boldsymbol{\eta} =& \left[
\begin {array}{cc}
 \eta_{11}& \eta_{12} \\
 \eta_{21}& \eta_{22} 
\end {array}
\right]
= \left[
\begin {array}{cc}
.8 & .2\\.2 & .8
\end {array}
\right] \\
\mbox{Dyad (i,j)} &  & y_{ij} \sim& F(\eta_{z_i,z_j}) \\
\mbox{Actor i's class} & & z_i \sim& \mbox{Categorical}(\pi)
\end{align*}

@@

\begin{frame}{Blockmodels for relational event data}
Idea:
  \begin{itemize}
  \item Assume each actor $i$ belongs to a latent class $z_i$.
  \item Model event dynamics between blocks.
  \end{itemize}

@@

Notation:
\begin{itemize}
\item $\boldsymbol{\beta}_{k,l}$: parameter vector for relational events from an actor in block $k$ to an actor in block $l$
\item  $\mathcal{A}_t$: history prior to time $t$
\item $\mathbf{z}$: vector of class assignments
\end{itemize}

@@

\begin{figure}

\centering 
\subfigure[Dynamic network data]
{
\begin{tikzpicture}[scale=.8,transform shape,thick]
  \node[vertex] (n1) at (0,0)  {1};
  \node[vertex] (n2) at (2,2)  {2};
  \node[vertex] (n3) at (4,0) {3};
  \node[vertex] (n4) at (2,-2) {4};
\draw[-stealth] (.4,.4) -- (1.6,1.6);
\draw[-stealth] (2.4,1.6) -- (3.6,.4);
\node at (.7,1.1) {$t_1$};
\node at (3.3,1.1) {$t_2$};
\end{tikzpicture}
}
\subfigure[Intensities for two dyads]
{
\begin{tikzpicture}[scale=1,transform shape,thick]
\draw[-stealth] (0,0) -- (0,3);
\draw[-stealth] (0,0) -- (5,0);

@@

% labels
\node at (-.5,2) {$\lambda_{1,2}$};
\node at (-.5,1) {$\lambda_{3,4}$};
\node at (1,-.3) {$t_1$};
\node at (3,-.3) {$t_2$};
\end{tikzpicture}
}
\end{figure}

@@

\begin{frame}{Synthetic example}
True parameters

Observed counts for each block.

Parameter estimates (using Gibbs sampling for $\mathbf{z}$ and slice sampling for $\boldsymbol{\beta}$).
\end{frame}

>>>>

\documentclass{beamer}
\usetheme{default}
\usepackage{tikz}
\usepackage{subfig}
\setlength{\unitlength}{\textwidth} 
\tikzstyle{vertex}=[auto=left,circle,fill=black!25,minimum size=20pt,inner sep=0pt]
\title[Blockmodels for relational events]%(optional, only for long titles)
{Stochastic blockmodels for relational events}
\subtitle{}
\author[DuBois, Christopher] % (optional, for multiple authors)
{Christopher~DuBois\inst{1} \and Carter~Butts\inst{2} \and Padhraic~Smyth\inst{3}}
\institute[] % (optional)
{
  \inst{1}%
Department of Statistics \\
University of California, Irvine
  \and
  \inst{2}%
Department of Sociology \\
University of California, Irvine
  \and
  \inst{2}%
Department of Computer Science \\
University of California, Irvine
}
\date[Sunbelt 2012] % (optional)
{Sunbelt 2012}
\subject{Informatik}
\begin{document}
\frame{\titlepage}

@@

\begin{align}
\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}}\exp\{ - (t_m - t_{m-1}) \lambda_{ij}(t_m | .) \}
\end{align}

\begin{itemize}
\item $M$: number of events
\item $t_m$: time of event $m$
\item $\mathcal{R}$: set of dyadic events
\end{itemize}

@@

\includegraphics[scale=.45]{../../figs/synthetic/unsorted}
\uncover<2->{\includegraphics[scale=.45]{../../figs/synthetic/mat}}
\uncover<3->{\includegraphics[scale=.45]{../../figs/synthetic/bm} }

\vspace{-.8cm}
\begin{align*}
\uncover<4->{
\mbox{Mixing rates} & & \boldsymbol{\eta} =& \left[
\begin {array}{cc}
 \eta_{11}& \eta_{12} \\
 \eta_{21}& \eta_{22} 
\end {array}
\right]
= \left[
\begin {array}{cc}
.8 & .2\\.2 & .8
\end {array}
\right] \\
}
\uncover<5->{
\mbox{Dyad (i,j)} &  & y_{ij} \sim& F(\eta_{z_i,z_j}) \\
}
\uncover<6->{
\mbox{Actor i's class} & & z_i \sim& \mbox{Categorical}(\pi)
}
\end{align*}

@@

\begin{frame}{Stochastic blockmodels for relational event data}
Idea:
  \begin{itemize}
  \item Assume each actor $i$ belongs to a latent class $z_i$.
  \item Model event dynamics between blocks.
  \end{itemize}

@@

\uncover<2>{Notation:
\begin{itemize}
\item $\boldsymbol{\beta}_{k,l}$: parameter vector for relational events from an actor in block $k$ to an actor in block $l$
\item  $\mathcal{A}_t$: history prior to time $t$
\item $\mathbf{z}$: vector of class assignments
\end{itemize}
}

\uncover<3>{
\vspace{-3cm}
Benefits:  
\begin{itemize}
\item Adjust for unobserved heterogeneity.
\item Obtain clusters of nodes who have similar interaction dynamic with other parts of the network.
\end{itemize}

}

@@

\begin{figure}
\centering 
%\subfigure[Dynamic network data]
%{
\begin{tikzpicture}[scale=.8,transform shape,thick]
  \node[vertex] (n1) at (0,0)  {1};
  \node[vertex] (n2) at (2,2)  {2};
  \node[vertex] (n3) at (4,0) {3};
  \node[vertex] (n4) at (2,-2) {4};
\draw[-stealth] (.4,.4) -- (1.6,1.6);
\draw[-stealth] (2.4,1.6) -- (3.6,.4);
\node at (.7,1.1) {$t_1$};
\node at (3.3,1.1) {$t_2$};
\end{tikzpicture}
% }
% \subfigure[Intensities for two dyads]
% {
\begin{tikzpicture}[scale=1,transform shape,thick]
\draw[-stealth] (0,0) -- (0,3);
\draw[-stealth] (0,0) -- (5,0);

@@

% labels
\node at (-.5,2) {$\lambda_{1,2}$};
\node at (-.5,1) {$\lambda_{3,4}$};
\node at (1,-.3) {$t_1$};
\node at (3,-.3) {$t_2$};
\end{tikzpicture}
%}
\end{figure}

@@

\begin{frame}{Synthetic example: $\beta$ and observed counts}
\begin{picture}(0.0,0.0)
   \put(-.1,-0.25){\includegraphics[width=0.7\textwidth]{../../figs/synthetic/params-true}}
   \put(.6,-0.3){\includegraphics[width=0.5\textwidth]{../../figs/synthetic/counts}}
\end{picture}
\end{frame}

\begin{frame}{Synthetic example: slice sampling for $\hat{\beta}$ and Gibbs on
    $\mathbf{z}$}
\begin{picture}(0.0,0.0)
   \put(-.1,-0.25){\includegraphics[width=0.7\textwidth]{../../figs/synthetic/params-estimates}}
   \put(.6,-0.3){\includegraphics[width=0.5\textwidth]{../../figs/synthetic/counts}}
\end{picture}

%\vspace{1in}
%From 100 samples after 50 burnin using Gibbs sampling for $\mathbf{z}$ and slice sampling for $\boldsymbol{\beta}$).
\end{frame}
